---
title: "Exploratory analysis on Activity monitoring data"
author: "Kishibe Rohan"
date: "14th July,2020"
output:
  html_document: default
  pdf_document: default
---

## Overview
It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the “quantified self” movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and there is a lack of statistical methods and software for processing and interpreting the data.

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The data for this assignment can be downloaded from the course web site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) 

The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as 𝙽𝙰) 
date: The date on which the measurement was taken in YYYY-MM-DD format 
interval: Identifier for the 5-minute interval in which measurement was taken 
The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset. 

## Loading and exploring the dataset 
Make sure the .csv file is in the same directory

```{r,echo=TRUE}
library(ggplot2)
activity <- read.csv("activity.csv")
```

```{r,results='hide'}
dim(activity)
names(activity)
head(activity)
str(activity)
```

##Histogram of the total number of steps taken each day

```{r,echo=TRUE}

#Fetching the required data 
dailySteps <- aggregate(steps ~ date,activity,sum,na.rm=TRUE)

#Plot the histogram

ggplot(dailySteps, aes(x =
                         steps)) +
    geom_histogram(fill = "blue", binwidth = 1000) +
    labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```
## Mean and median of daily total steps
```{r}
meanDailySteps <- mean(dailySteps$steps,na.rm = TRUE)
meanDailySteps

```

```{r}
medianDailySteps <- median(dailySteps$steps,na.rm=TRUE)
medianDailySteps

```

## Average daily activity pattern

1. Make a time series plot (i.e. 𝚝𝚢𝚙𝚎 = "𝚕") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}
stepsPerInterval <- aggregate(steps~interval,data=activity,mean,na.rm=TRUE)

ggplot(stepsPerInterval,aes(x=interval,y=steps))+geom_line(color="blue",size = 1) + labs(title = "Average Daily Steps", x = "Interval", y = "Steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
maxStepsInterval <- stepsPerInterval[which.max(stepsPerInterval$steps),]$interval

maxStepsInterval
```

## Imputing missing values
1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with 𝙽𝙰s)

```{r}
totalMissing <- sum(is.na(activity$steps))
totalMissing
```

2.Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. 

 Strategy used : Fill in all the missing values in the dataset with the mean per interval.
 
```{r}
 
 meanStepsPerInterval <- function(interval)
 {
     stepsPerInterval[stepsPerInterval$interval==interval,]$steps
 }

```

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.


```{r}
 tidyData <- activity
 for(i in 1:nrow(tidyData))
 {
 if(is.na(tidyData[i,]$steps))
 {
   tidyData[i,]$steps <- meanStepsPerInterval(tidyData[i,]$interval)
 }
 }



```

 4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day
 
```{r}

stepsPerDayNoNA <- aggregate(steps ~ date, tidyData, sum)



 ggplot(stepsPerDayNoNA, aes(x =
                         steps)) +
    geom_histogram(fill = "blue", binwidth = 1000) +
    labs(title = "Daily Steps", x = "Steps", y = "Frequency")
```
  
The mean didn’t change after the replacements of NAs, the median changed about 0.1% of the original value.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r}
tidyData$date <- as.Date(strptime(tidyData$date,format ="%Y-%m-%d" ))
tidyData$day <- weekdays(tidyData$date)

for (i in 1:nrow(tidyData)) {
    if (tidyData[i,]$day %in%
        c("Saturday","Sunday")) {
        tidyData[i,]$day<-"weekend"
    }
    else{
        tidyData[i,]$day<-"weekday"
    }
}

stepsByDay <- aggregate(tidyData$steps ~ tidyData$interval + tidyData$day,tidyData,mean)

```

2. Make a panel plot containing a time series plot (i.e. 𝚝𝚢𝚙𝚎 = “𝚕”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r}
library(lattice)
names(stepsByDay) <- c("interval","day","steps")
xyplot(steps~interval|day,stepsByDay,type="l",layout=c(1,2),xlab="Interval",ylab="Steps")

```

